```{r setup, include=F}
library(ComplexHeatmap)
library(pheatmap)
library(eulerr)
library(tidyverse)
source("scripts/helpers.R")

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 7,
  fig.show = "hold",
  out.width = "50%"
)
```

```{r read data, include=F}
source("scripts/main-data.R")
source("scripts/compare-genelists-function.R")
```

clonalQ1: RN: For  wt_NS how many DEGs are different (≥2/≤0.5 -fold, P ≤ 0.05) compared to: a) sham_NS and b) to KO_NS?

AG: I'm going to take this to mean significantly different genes in sham and KO since I did the NS-only dea with wt as reference level.

```{r clonalQ1a}
sham_ns_74 = dea_ns %>%
  filter(group == "sham_naive", sig != "ns") %>%
  pull(Gene)

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

genelists = list(
  sham_ns_74 = sham_ns_74,
  wt_up_738 = wt_up_738
)

plot_gene_sets(
  genelists,
  mat_ns,
  mat_cols = 1:8,
  plotType = "euler",
  heatmapTitle = "log2tpm_NS"
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = 1:8,
  plotType = "euler",
  heatmapTitle = "log2fold"
)



cleanup_objects()
```

```{r clonalQ1b}
ko_ns_682 = dea_ns %>%
  filter(group == "TAK1KO_naive", sig != "ns") %>%
  pull(Gene)

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

genelists = list(
  ko_ns_682 = ko_ns_682,
  wt_up_738 = wt_up_738
)

plot_gene_sets(
  genelists,
  mat_ns,
  mat_cols = c(1:4, 9:12),
  plotType = "euler",
  heatmapTitle = "log2tpm_NS"
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = c(1:4, 9:12),
  plotType = "euler",
  heatmapTitle = "log2fold"
)

# paste(
#   intersect(ko_ns_682, wt_up_738),
#   collapse = ", "
# )


cleanup_objects()
```



clonalQ2: RN: For  wt_IL-1β how many DEGs are different (≥2/≤0.5 -fold, P ≤ 0.05) compared to: a) sham IL-1β and b) to KO IL-1β?

AG: I'm going to take this to mean significantly different genes in sham and KO since I did the IL1B-only dea with wt as reference level.


```{r clonalQ2a}
sham_il1b_80 = dea_il1b %>%
  filter(group == "sham_naive", sig != "ns") %>%
  pull(Gene)

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

genelists = list(
  sham_il1b_80 = sham_il1b_80,
  wt_up_738 = wt_up_738
)

plot_gene_sets(
  genelists,
  mat_il1b,
  mat_cols = 1:8,
  plotType = "euler",
  heatmapTitle = "log2tpm_IL1B"
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = 1:8,
  plotType = "euler",
  heatmapTitle = "log2fold"
)



cleanup_objects()
```

```{r clonalQ2b}
ko_il1b_1484 = dea_il1b %>%
  filter(group == "TAK1KO_naive", sig != "ns") %>%
  pull(Gene)

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

genelists = list(
  ko_il1b_1484 = ko_il1b_1484,
  wt_up_738 = wt_up_738
)

plot_gene_sets(
  genelists,
  mat_il1b,
  mat_cols = c(1:4, 9:12),
  plotType = "euler",
  heatmapTitle = "log2tpm_IL1B"
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = c(1:4, 9:12),
  plotType = "euler",
  heatmapTitle = "log2fold"
)

cleanup_objects()
```



clonalQ3: RN: I presume that most of the differences will be clonal and therefore common to both the above? True/false – let’s see?

```{r clonalQ3a, fig.height=10}
sham_ns_74 = dea_ns %>%
  filter(group == "sham_naive", sig != "ns") %>%
  pull(Gene)

sham_il1b_80 = dea_il1b %>%
  filter(group == "sham_naive", sig != "ns") %>%
  pull(Gene)

genelists = list(
  sham_ns_74 = sham_ns_74,
  sham_il1b_80 = sham_il1b_80
)

# plot_gene_sets(
#   genelists,
#   mat_ns,
#   mat_cols = 1:8,
#   plotType = "euler",
#   heatmapTitle = "log2tpm_NS"
# )

plot_gene_sets(
  genelists,
  mat,
  mat_cols = 1:8,
  plotType = "euler",
  heatmapTitle = "log2fold",
  heatmapRowNames = TRUE
)

# ===== drop log2fold criteria =====

union_110 = unique(c(sham_ns_74, sham_il1b_80))

sham_ns_l2f_dropped = dea_ns %>%
  filter(group == "sham_naive", FDR <= 0.05, Gene %in% union_110) %>%
  pull(Gene)

sham_il1b_l2f_dropped = dea_il1b %>%
  filter(group == "sham_naive", FDR <= 0.05, Gene %in% union_110) %>%
  pull(Gene)

genelists = list(
  sham_ns_l2f_dropped = sham_ns_l2f_dropped,
  sham_il1b_l2f_dropped = sham_il1b_l2f_dropped
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = 1:8,
  plotType = "euler",
  heatmapTitle = "log2fold",
  heatmapRowNames = TRUE
)



cleanup_objects()

```

"cxcl5-like genes"

```{r clonalQ3b, fig.width=7}
ko_ns_682 = dea_ns %>%
  filter(group == "TAK1KO_naive", sig != "ns") %>%
  pull(Gene)

ko_il1b_1484 = dea_il1b %>%
  filter(group == "TAK1KO_naive", sig != "ns") %>%
  pull(Gene)

genelists = list(
  ko_ns_682 = ko_ns_682,
  ko_il1b_1484 = ko_il1b_1484
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = c(1:4, 9:12),
  plotType = "euler",
  heatmapTitle = "log2fold"
)

# cxcl5-like genes are invisible to fold-based analysis, but are knocked out
# and restored with TAK1 KO/rescue. 

# they exist at the intersect between ko_ns_682 and ko_il1b_1484.
# there are 489 suspected cxcl5-like genes.

cxcl5_like_489 = intersect(ko_ns_682, ko_il1b_1484)

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

genelists = list(
  cxcl5_like_489 = cxcl5_like_489,
  wt_up_738 = wt_up_738
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = c(1:4, 9:12),
  plotType = "euler",
  heatmapTitle = "log2fold"
)

cxcl5_like_up = intersect(cxcl5_like_489, wt_up_738)

paste(cxcl5_like_up, collapse = ", ")

Heatmap(
  mat_combined[cxcl5_like_up, ],
  cluster_columns = F,
  show_column_names = F,
  column_title_rot = 90, 
  
  show_row_names = T,
  row_names_gp = gpar(cex = 0.8),
  
  col = generic_l2f_heatmap_colors(mat_combined[cxcl5_like_up, ]),
  
  name = "log2tpm",
  
  column_split = factor(
    c(
      rep("WT NS", 4),
      rep("WT IL1B", 4),
      rep("sham NS", 4),
      rep("sham IL1B", 4),
      rep("KO NS", 4),
      rep("KO IL1B", 4),
      rep("lip NS", 4),
      rep("lip IL1B", 4),
      rep("nat NS", 4),
      rep("nat IL1B", 4),
      rep("D175A NS", 4),
      rep("D175A IL1B", 4)
    ),
    levels = c(
      "WT NS",
      "WT IL1B",
      "sham NS",
      "sham IL1B",
      "KO NS",
      "KO IL1B",
      "lip NS",
      "lip IL1B",
      "nat NS",
      "nat IL1B",
      "D175A NS",
      "D175A IL1B"
    )
  )
)

Heatmap(
  mat[cxcl5_like_up, ],
  cluster_columns = F,
  show_column_names = T,
  column_names_rot = 45, 
  show_row_names = T,
  row_names_gp = gpar(cex = 0.5),
  col = generic_l2f_heatmap_colors(mat_combined[cxcl5_like_up, ]),
  name = "log2fold"
)


cleanup_objects()

```




clonalQ4 and 5: 

For DEGs that are differentially expressed in 1 &/or 2, how many get rescued by native TAK1 smRNA?

For DEGs that are differentially expressed in 1 &/or 2, how many get rescued by D175A TAK1 smRNA?

```{r, fig.width=7}
sham_ns_74 = dea_ns %>%
  filter(group == "sham_naive", sig != "ns") %>%
  pull(Gene)

ko_ns_682 = dea_ns %>%
  filter(group == "TAK1KO_naive", sig != "ns") %>%
  pull(Gene)

nat_ns_861 = dea_ns %>%
  filter(group == "TAK1KO_TAK1-smRNA", sig == "up") %>%
  pull(Gene)

mutant_ns_336 = dea_ns %>%
  filter(group == "TAK1KO_mTAK1-smRNA", sig == "up") %>%
  pull(Gene)



genelists = list(
  sham_ns_74 = sham_ns_74,
  ko_ns_682 = ko_ns_682,
  nat_ns_861 = nat_ns_861
)

plot_gene_sets(
  genelists,
  mat,
  plotType = "euler"
)



genelists = list(
  sham_ns_74 = sham_ns_74,
  ko_ns_682 = ko_ns_682,
  mutant_ns_336 = mutant_ns_336
)

plot_gene_sets(
  genelists,
  mat,
  plotType = "euler"
)


cleanup_objects()

```

