```{r setup, include=F}
library(ComplexHeatmap)
library(pheatmap)
library(eulerr)
library(tidyverse)
source("scripts/helpers.R")

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 7,
  fig.show = "hold",
  out.width = "50%"
)
```

```{r read data, include=F}
source("scripts/main-data.R")
source("scripts/compare-genelists-function.R")
```

738 DEGs upregulated by IL-1β (≥2-fold, P ≤ 0.05) 




Q1a: RE sham - How many DEGs are induced (≥2-fold, P ≤ 0.05) in the sham? How many are significantly different (≥2/≤0.5 -fold, P ≤ 0.05) from wt?

```{r}
# ===== with log2fold and p value criteria =====

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

sham_up_751 = dea %>% 
  filter(condition == "sham", sig == "up") %>% 
  pull(Gene)

genelists = list(
  wt_up_738 = wt_up_738,
  sham_up_751 = sham_up_751
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = 1:8
)

# ===== calculate correlations ======

for(c in c("01", "11", "10")){
  
  goi = plot_gene_sets(
    genelists,
    mat,
    output = "codes"
  ) %>%
    filter(code == c) %>%
    pull(Gene)
  
  df = dea %>%
    filter(group %in% c("WT_naive", "sham_naive"), Gene %in% goi) %>%
    select(Gene, group, log2fold) %>%
    pivot_wider(names_from = "group", values_from = "log2fold")
  
  cor = cor.test(df$WT_naive, df$sham_naive, method = "pearson")$estimate[[1]]
  
  print(paste0("code:", c, " pearson= ", cor, " (", length(goi), " genes)"))
  
  rm(c, goi, df, cor)
  
}


cleanup_objects()
```

Q1a.1: effect of just going with the P-value (i.e., drop fold criteria?) Does it collapse into a single group.

```{r}
# ===== use log2fold + pvalue to select, but then drop log2fold =====

# establish 844 gene pool

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

sham_up_751 = dea %>% 
  filter(condition == "sham", sig == "up") %>% 
  pull(Gene)

union_844 = unique(c(wt_up_738, sham_up_751))

# pull pool genes, then drop to pvalue criteria only only

wt_up_l2f_dropped = dea %>%
  filter(condition == "WT", Gene %in% union_844) %>%
  filter(FDR <= 0.05, log2fold > 0) %>%
  pull(Gene)

sham_up_l2f_dropped = dea %>%
  filter(condition == "sham", Gene %in% union_844) %>%
  filter(FDR <= 0.05, log2fold > 0) %>%
  pull(Gene)

genelists = list(
  wt_up_l2f_dropped = wt_up_l2f_dropped,
  sham_up_l2f_dropped = sham_up_l2f_dropped
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = 1:8
)

# ===== calculate correlations ======

for(c in c("01", "11", "10")){
  
  goi = plot_gene_sets(
    genelists,
    mat,
    output = "codes"
  ) %>%
    filter(code == c) %>%
    pull(Gene)
  
  df = dea %>%
    filter(group %in% c("WT_naive", "sham_naive"), Gene %in% goi) %>%
    select(Gene, group, log2fold) %>%
    pivot_wider(names_from = "group", values_from = "log2fold")
  
  cor = cor.test(df$WT_naive, df$sham_naive, method = "pearson")$estimate[[1]]
  
  print(paste0("code:", c, " pearson= ", cor, " (", length(goi), " genes)"))
  
  rm(c, goi, df, cor)
  
}

cleanup_objects()

```


Q1b: In how many genes is the IL1B response affected by sham (P<= 0.05)?

```{r fig.height=3, fig.width=4}
sham_il1b_interaction = read_tsv("data/sham_IL1B_interaction_test.txt")

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

# sham_up_751 = dea %>% 
#   filter(condition == "sham", sig == "up") %>% 
#   pull(Gene)
# 
# union_844 = unique(c(wt_up_738, sham_up_751))

df = as.data.frame(wt_up_738, nm = "Gene") %>%
  left_join(
    sham_il1b_interaction %>% 
      select(Gene, log2diff, FDR)
    )

hist(df$log2diff, breaks = 100)
summary(df$log2diff)

hist(df$FDR, breaks = 100)
summary(df$FDR)

df %>%
  mutate(
    direction = case_when(
      log2diff == 0  ~ "0 or NA log2diff",
      log2diff <= 0 ~ "-ve log2diff",
      log2diff >= 0  ~ "+ve log2diff"
    )
  ) %>%
  count(direction) %>%
  mutate(percent = 100*n/sum(n))

df %>%
  mutate(sig = if_else(FDR <= 0.05, "FDR<=0.05", "ns")) %>%
  count(sig) %>%
  mutate(percent = 100*n/sum(n))

cleanup_objects()
```




Q2a: RE KO - How many DEGs are induced (≥2-fold, P ≤ 0.05) in the KO? How many are significantly different (≥2/≤0.5 -fold, P ≤ 0.05) from wt?

```{r}
wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

ko_up_71 = dea %>% 
  filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>% 
  pull(Gene)

genelists = list(
  wt_up_738 = wt_up_738,
  ko_up_71 = ko_up_71
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = c(1:4, 9:12)
)

# intersect(wt_up_738, ko_up_71) %>% paste(collapse = ", ")
# 
# setdiff(ko_up_71, wt_up_738)

# ===== calculate correlations ======

for(c in c("01", "11", "10")){
  
  goi = plot_gene_sets(
    genelists,
    mat,
    output = "codes"
  ) %>%
    filter(code == c) %>%
    pull(Gene)
  
  df = dea %>%
    filter(group %in% c("WT_naive", "TAK1KO_naive"), Gene %in% goi) %>%
    select(Gene, group, log2fold) %>%
    pivot_wider(names_from = "group", values_from = "log2fold")
  
  cor = cor.test(df$WT_naive, df$TAK1KO_naive, method = "pearson")$estimate[[1]]
  
  print(paste0("code:", c, " pearson= ", cor, " (", length(goi), " genes)"))
  
  rm(c, goi, df, cor)
  
}

cleanup_objects()
```

Q2a.1: effect of just going with the P-value (i.e., drop fold criteria?) Does it collapse into a single group.

```{r}
# ===== use log2fold + pvalue to select, but then drop log2fold =====

# establish 741 gene pool

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

ko_up_71 = dea %>% 
  filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>% 
  pull(Gene)

union_741 = unique(c(wt_up_738, ko_up_71))

# pull pool genes, then drop to pvalue criteria only only

wt_up_l2f_dropped = dea %>%
  filter(condition == "WT", Gene %in% union_741) %>%
  filter(FDR <= 0.05, log2fold > 0) %>%
  pull(Gene)

ko_up_l2f_dropped = dea %>%
  filter(group == "TAK1KO_naive", Gene %in% union_741) %>%
  filter(FDR <= 0.05, log2fold > 0) %>%
  pull(Gene)

genelists = list(
  wt_up_l2f_dropped = wt_up_l2f_dropped,
  ko_up_l2f_dropped = ko_up_l2f_dropped
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = 1:8
)

# ===== calculate correlations ======

for(c in c("01", "11", "10")){
  
  goi = plot_gene_sets(
    genelists,
    mat,
    output = "codes"
  ) %>%
    filter(code == c) %>%
    pull(Gene)
  
  df = dea %>%
    filter(group %in% c("WT_naive", "sham_naive"), Gene %in% goi) %>%
    select(Gene, group, log2fold) %>%
    pivot_wider(names_from = "group", values_from = "log2fold")
  
  cor = cor.test(df$WT_naive, df$sham_naive, method = "pearson")$estimate[[1]]
  
  print(paste0("code:", c, " pearson= ", cor, " (", length(goi), " genes)"))
  
  rm(c, goi, df, cor)
  
}

cleanup_objects()

```

Q2b: In how many genes is the IL1B response affected by TAK1KO (P<= 0.05)?

```{r fig.height=3, fig.width=4}
tak1ko_il1b_interaction = read_tsv("data/KO_IL1B_interaction_test.txt")

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

# ko_up_71 = dea %>% 
#   filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>% 
#   pull(Gene)

# union_741 = unique(c(wt_up_738, ko_up_71))

df = as.data.frame(wt_up_738, nm = "Gene") %>%
  left_join(
    tak1ko_il1b_interaction %>% 
      select(Gene, log2diff, FDR)
    )

hist(df$log2diff, breaks = 100)
summary(df$log2diff)


hist(df$FDR, breaks = 100)
summary(df$FDR)

df %>%
  mutate(
    direction = case_when(
      log2diff == 0  ~ "0 or NA log2diff",
      log2diff <= 0 ~ "-ve log2diff",
      log2diff >= 0  ~ "+ve log2diff"
    )
  ) %>%
  count(direction) %>%
  mutate(percent = 100*n/sum(n))

df %>%
  mutate(sig = if_else(FDR <= 0.05, "FDR<=0.05", "ns")) %>%
  count(sig) %>%
  mutate(percent = 100*n/sum(n))

cleanup_objects()
```






RN, Q3: RE KO+lipid – Does the lipid change anything? So: How many DEGs are also induced (≥2-fold, P ≤ 0.05) in the KO+lipid? How many are significantly different (≥2/≤0.5 -fold, P ≤ 0.05) from KO? Ideally there are similar numbers to the KO and that very few DEGs change.

```{r}
ko_up_71 = dea %>% 
  filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>% 
  pull(Gene)

lipid_up_68 = dea %>% 
  filter(condition == "TAK1KO", rescue == "lipid", sig == "up") %>% 
  pull(Gene)

genelists = list(
  ko_up_71 = ko_up_71,
  lipid_up_68 = lipid_up_68
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = c(9:16)
)

# ===== calculate correlations ======

for(c in c("01", "11", "10")){
  
  goi = plot_gene_sets(
    genelists,
    mat,
    output = "codes"
  ) %>%
    filter(code == c) %>%
    pull(Gene)
  
  df = dea %>%
    filter(group %in% c("TAK1KO_naive", "TAK1KO_lipid"), Gene %in% goi) %>%
    select(Gene, group, log2fold) %>%
    pivot_wider(names_from = "group", values_from = "log2fold")
  
  cor = cor.test(df$TAK1KO_naive, df$TAK1KO_lipid, method = "pearson")$estimate[[1]]
  
  print(paste0("code:", c, " pearson= ", cor, " (", length(goi), " genes)"))
  
  rm(c, goi, df, cor)
  
}

cleanup_objects()
```

Q3.1: effect of just going with the P-value (i.e., drop fold criteria?) Does it collapse into a single group.

```{r}
# ===== use log2fold + pvalue to select, but then drop log2fold =====

# establish 78 gene pool

ko_up_71 = dea %>% 
  filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>% 
  pull(Gene)

lipid_up_68 = dea %>% 
  filter(condition == "TAK1KO", rescue == "lipid", sig == "up") %>% 
  pull(Gene)

union_78 = unique(c(ko_up_71, lipid_up_68))

# pull pool genes, then drop to pvalue criteria only only

ko_up_l2f_dropped = dea %>%
  filter(group == "TAK1KO_naive", Gene %in% union_78) %>%
  filter(FDR <= 0.05, log2fold > 0) %>%
  pull(Gene)

lipid_up_l2f_dropped = dea %>%
  filter(group == "TAK1KO_lipid", Gene %in% union_78) %>%
  filter(FDR <= 0.05, log2fold > 0) %>%
  pull(Gene)

genelists = list(
  ko_up_l2f_dropped = ko_up_l2f_dropped,
  lipid_up_l2f_dropped = lipid_up_l2f_dropped
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = 1:8
)

# ===== calculate correlations ======

for(c in c("01", "11", "10")){
  
  goi = plot_gene_sets(
    genelists,
    mat,
    output = "codes"
  ) %>%
    filter(code == c) %>%
    pull(Gene)
  
  df = dea %>%
    filter(group %in% c("TAK1KO_naive", "TAK1KO_lipid"), Gene %in% goi) %>%
    select(Gene, group, log2fold) %>%
    pivot_wider(names_from = "group", values_from = "log2fold")
  
  cor = cor.test(df$TAK1KO_naive, df$TAK1KO_lipid, method = "pearson")$estimate[[1]]
  
  print(paste0("code:", c, " pearson= ", cor, " (", length(goi), " genes)"))
  
  rm(c, goi, df, cor)
  
}

cleanup_objects()

```










Q4a: RE native TAK1 smRNA in the KO cells: a) How many DEGs are induced (≥2-fold, P ≤ 0.05) in the TAK1 KO + TAK1 smRNA? b) Compared to TAK1 KO, how many DEGs are significantly different (≥2/≤0.5 -fold, P ≤ 0.05)? Note: could be legit to do this vs KO+lipid?

```{r}
ko_up_71 = dea %>% 
  filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>% 
  pull(Gene)

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

rescue_up_540 = dea %>% 
  filter(condition == "TAK1KO", rescue == "TAK1-smRNA", sig == "up") %>% 
  pull(Gene)

genelists = list(
  ko_up_71 = ko_up_71,
  wt_up_738= wt_up_738,
  rescue_up_540 = rescue_up_540
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = c(1:4, 9:12, 17:20),
  plotType = "euler"
)

# intersect(ko_up_71, rescue_up_540) %>% paste(collapse = ", ")

cleanup_objects()
```

Q4a.1: take the above Venn and drop the log2fold criteria? See if it collapses.

```{r}
# establish gene pool

ko_up_71 = dea %>% 
  filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>% 
  pull(Gene)

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

rescue_up_540 = dea %>% 
  filter(condition == "TAK1KO", rescue == "TAK1-smRNA", sig == "up") %>% 
  pull(Gene)

union_857 = unique(c(
  ko_up_71,
  wt_up_738,
  rescue_up_540
))

# re-establish gene pools with l2f threshold dropped

ko_up_l2f_dropped = dea %>%
  filter(
    condition == "TAK1KO", 
    rescue == "naive", 
    FDR <= 0.05,
    Gene %in% union_857
    ) %>%
  pull(Gene)

wt_up_l2f_dropped = dea %>%
  filter(
    condition == "WT", 
    FDR <= 0.05,
    Gene %in% union_857
    ) %>%
  pull(Gene)

rescue_up_l2f_dropped = dea %>%
  filter(
    condition == "TAK1KO",
    rescue == "TAK1-smRNA",
    FDR <= 0.05,
    Gene %in% union_857
  ) %>%
  pull(Gene)

genelists = list(
  ko_up_l2f_dropped = ko_up_l2f_dropped,
  wt_up_l2f_dropped = wt_up_l2f_dropped,
  rescue_up_l2f_dropped = rescue_up_l2f_dropped
)

plot_gene_sets(
  genelists,
  mat,
  mat_cols = c(1:4, 9:12, 17:20),
  plotType = "euler"
)


```


Q4b: effect of introducing native TAK1 smRNA to TAK1KO?

```{r, fig.height=3, fig.width=4}
rescue_interaction = read_tsv("data/nativeRescue_interaction_test.txt")

# ko_up_71 = dea %>% 
#   filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>% 
#   pull(Gene)

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

# rescue_up_540 = dea %>% 
#   filter(condition == "TAK1KO", rescue == "TAK1-smRNA", sig == "up") %>% 
#   pull(Gene)

# union_857 = unique(c(
#   ko_up_71,
#   wt_up_738,
#   rescue_up_540
# ))

df = rescue_interaction %>%
  filter(Gene %in% wt_up_738)

hist(df$log2diff, breaks = 100)
summary(df$log2diff)


hist(df$FDR, breaks = 100)
summary(df$FDR)


df %>%
  mutate(
    direction = case_when(
      log2diff == 0  ~ "0 or NA log2diff",
      log2diff <= 0 ~ "-ve log2diff",
      log2diff >= 0  ~ "+ve log2diff"
    )
  ) %>%
  count(direction) %>%
  mutate(percent = 100*n/sum(n))

df %>%
  mutate(sig = if_else(FDR <= 0.05, "FDR<=0.05", "ns")) %>%
  count(sig) %>%
  mutate(percent = 100*n/sum(n))



cleanup_objects()
```


Q4c: compare WT and rescued IL1B response?

```{r, fig.height=3, fig.width=4}
wt_rescue_interaction = read_tsv("data/wt_vs_rescued_interaction_test.txt")

# establish gene pool

# ko_up_71 = dea %>%
#   filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>%
#   pull(Gene)

wt_up_738 = dea %>% 
  filter(condition == "WT", sig == "up") %>% 
  pull(Gene)

# rescue_up_540 = dea %>% 
#   filter(condition == "TAK1KO", rescue == "TAK1-smRNA", sig == "up") %>% 
#   pull(Gene)

# union_857 = unique(c(
#   ko_up_71,
#   wt_up_738,
#   rescue_up_540
# ))

df = as.data.frame(wt_up_738, nm = "Gene") %>%
  left_join(
    wt_rescue_interaction %>% 
      select(Gene, log2diff, FDR)
    )

hist(df$log2diff, breaks = 100)
summary(df$log2diff)

hist(df$FDR, breaks = 100)
summary(df$FDR)

df %>%
  mutate(sig = if_else(FDR <= 0.05, "FDR<=0.05", "ns")) %>%
  count(sig) %>%
  mutate(percent = 100*n/sum(n))

df %>%
  mutate(
    direction = case_when(
      log2diff == 0  ~ "0 or NA log2diff",
      log2diff <= 0 ~ "-ve log2diff",
      log2diff >= 0  ~ "+ve log2diff"
    )
  ) %>%
  count(direction) %>%
  mutate(percent = 100*n/sum(n))

cleanup_objects()
```


<!-- Q5: RE D175A TAK1 smRNA in the KO cells: a) How many DEGs are induced (≥2-fold, P ≤ 0.05) in the TAK1 KO + TAK1 smRNA? b) Compared to TAK1 KO, how many DEGs are significantly different (≥2/≤0.5 -fold, P ≤ 0.05)? Note: could be legit to do this vs KO+lipid? -->

<!-- ```{r} -->
<!-- ko_up_71 = dea %>%  -->
<!--   filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>%  -->
<!--   pull(Gene) -->

<!-- mutant_up_139 = dea %>%  -->
<!--   filter(condition == "TAK1KO", rescue == "mTAK1-smRNA", sig == "up") %>%  -->
<!--   pull(Gene) -->

<!-- genelists = list( -->
<!--   ko_up_71 = ko_up_71, -->
<!--   mutant_up_139 = mutant_up_139 -->
<!-- ) -->

<!-- plot_gene_sets( -->
<!--   genelists, -->
<!--   mat, -->
<!--   mat_cols = c(9:12, 21:24) -->
<!-- ) -->

<!-- intersect(ko_up_71, mutant_up_139) %>% paste(collapse = ", ") -->

<!-- cleanup_objects() -->
<!-- ``` -->

<!-- ```{r fig.height=7, fig.width=7} -->
<!-- wt_up_738 = dea %>%  -->
<!--   filter(condition == "WT", sig == "up") %>%  -->
<!--   pull(Gene) -->

<!-- mutant_up_139 = dea %>%  -->
<!--   filter(condition == "TAK1KO", rescue == "mTAK1-smRNA", sig == "up") %>%  -->
<!--   pull(Gene) -->

<!-- genelists = list( -->
<!--   wt_up_738 = wt_up_738, -->
<!--   mutant_up_139 = mutant_up_139 -->
<!-- ) -->

<!-- plot_gene_sets( -->
<!--   genelists, -->
<!--   mat, -->
<!--   mat_cols = c(1:4, 21:24) -->
<!-- ) -->


<!-- cor = tpm %>%  -->
<!--   filter(group %in% c("WT_naive", "TAK1KO_mTAK1-smRNA")) %>% -->
<!--   filter(Gene %in% unique(c(wt_up_738, mutant_up_139))) %>% -->
<!--   select(Gene, group, log2fold) %>% -->
<!--   group_by(Gene, group) %>% -->
<!--   summarize(log2fold = mean(log2fold), .groups = "drop") %>% -->
<!--   pivot_wider(names_from = "group", values_from = "log2fold") -->

<!-- cor.test(cor$`TAK1KO_mTAK1-smRNA`, cor$WT_naive, method = "pearson") -->

<!-- ggplot( -->
<!--   cor, -->
<!--   aes(x = WT_naive, y = `TAK1KO_mTAK1-smRNA`) -->
<!-- ) + -->
<!--   geom_point() + -->
<!--   geom_smooth(method = "lm") + -->
<!--   theme(aspect.ratio = 1) -->

<!-- # cleanup_objects() -->
<!-- ``` -->

<!-- ```{r a confident rescue list} -->
<!-- wt_up_738 = dea %>%  -->
<!--   filter(condition == "WT", sig == "up") %>%  -->
<!--   pull(Gene) -->

<!-- ko_up_71 = dea %>%  -->
<!--   filter(condition == "TAK1KO", rescue == "naive", sig == "up") %>%  -->
<!--   pull(Gene) -->

<!-- rescue_up_540 = dea %>%  -->
<!--   filter(condition == "TAK1KO", rescue == "TAK1-smRNA", sig == "up") %>%  -->
<!--   pull(Gene) -->

<!-- mutant_up_139 = dea %>%  -->
<!--   filter(condition == "TAK1KO", rescue == "mTAK1-smRNA", sig == "up") %>%  -->
<!--   pull(Gene) -->


<!-- confident_rescue = setdiff( -->
<!--   intersect(wt_up_738, rescue_up_540), -->
<!--   # union(ko_up_71, mutant_up_139) -->
<!--   # ko_up_71 -->
<!--   mutant_up_139 -->
<!-- ) -->

<!-- paste0(confident_rescue, collapse = ", ") -->
<!-- ``` -->




